---
title: "The Mantel-Fleiss Criterion"
date: "2026-09-09"
output:
    rmarkdown::html_document:
        theme: "spacelab"
        highlight: "kate"
        toc: true
        toc_float: true
vignette: >
    %\VignetteIndexEntry{The Mantel-Fleiss Criterion}
    %\VignetteEngine{knitr::rmarkdown}
    %\VignetteEncoding{UTF-8}
bibliography: ../inst/REFERENCES.bib
editor_options:
    markdown:
        wrap: 72
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(tern)
```

## Introduction

When comparing a binary response between two groups while adjusting for a
stratification variable, the Cochran-Mantel-Haenszel (CMH) test is a common
choice. Like other large-sample procedures, the CMH test relies on an
asymptotic (chi-square) approximation, which can be unreliable when the
stratified $2 \times 2$ tables are sparse. In those situations an exact method
is preferable.

The **Mantel-Fleiss criterion** [@MantelFleiss1980] is a simple, quick check
that tells you whether the sample is large enough for the asymptotic CMH
approximation to be trustworthy. The `mantel_fleiss_crit()` function in `tern`
evaluates this criterion for a stratified $2 \times 2$ contingency table and
returns whether it is satisfied. You can use the result to decide, in a data
driven way, whether to run the CMH test or fall back to an exact procedure.

`mantel_fleiss_crit()` is a standalone utility: it does not perform any test
itself. It is meant to be used alongside the proportion functions in `tern`
(such as `prop_diff_cmh()`, `prop_cmh()`, `prop_diff_uncond_exact()`, and
`prop_fisher()`) when writing custom analysis functions.

## The criterion

Consider a stratified $2 \times 2$ table where $h$ indexes the strata. Within
stratum $h$, write the cell and margin counts as

$$
\begin{array}{c|cc|c}
& \text{Response} & \text{No response} & \text{Row total} \\
\hline
\text{Group 1} & n_{11h} & n_{12h} & n_{1 \cdot h} \\
\text{Group 2} & n_{21h} & n_{22h} & n_{2 \cdot h} \\
\hline
\text{Column total} & n_{\cdot 1 h} & n_{\cdot 2 h} & n_{h}
\end{array}
$$

Under the hypothesis of no association between group and response, the expected
count in cell $(1, 1)$ of stratum $h$ is

$$
m_{11h} = \frac{n_{1 \cdot h}\, n_{\cdot 1 h}}{n_{h}} .
$$

Given the fixed margins, the observed count $n_{11h}$ can range between the
bounds

$$
(n_{11h})_L = \max(0,\ n_{1 \cdot h} - n_{\cdot 2 h}), \qquad
(n_{11h})_U = \min(n_{\cdot 1 h},\ n_{1 \cdot h}) .
$$

The Mantel-Fleiss statistic aggregates these quantities across the non-empty
strata:

$$
MF = \min \left(
\left[ \sum_h m_{11h} - \sum_h (n_{11h})_L \right],\
\left[ \sum_h (n_{11h})_U - \sum_h m_{11h} \right]
\right) .
$$

The criterion is considered **satisfied** when $MF \ge$ `threshold`. The
default `threshold = 5` corresponds to the rule proposed by @MantelFleiss1980:
when $MF \ge 5$, the asymptotic CMH approximation is generally adequate.

## Basic usage

`mantel_fleiss_crit()` expects a three-dimensional contingency table (an
`array`) whose first two dimensions are the group and response (each with two
levels, in either order) and whose third dimension is the stratum. The
`table()` function produces exactly this layout:

```{r}
set.seed(123)
n <- 80

grp <- factor(sample(c("Active", "Control"), n, replace = TRUE))
rsp <- sample(c(TRUE, FALSE), n, replace = TRUE)
strata1 <- factor(sample(c("A", "B"), n, replace = TRUE))
strata2 <- factor(sample(c("x", "y"), n, replace = TRUE))
strata <- interaction(strata1, strata2)

tbl <- table(grp, rsp, strata)
tbl
```

Passing the table to `mantel_fleiss_crit()` returns a single logical value:

```{r}
mantel_fleiss_crit(tbl)
```

To see the underlying statistic, set `include_value = TRUE`. The Mantel-Fleiss
value is then attached to the result as a `"value"` attribute:

```{r}
mantel_fleiss_crit(tbl, include_value = TRUE)
```

The `threshold` argument controls how large the statistic must be for the
criterion to hold. Raising it makes the criterion more conservative:

```{r}
mantel_fleiss_crit(tbl, threshold = 100, include_value = TRUE)
```

## Choosing a test based on the criterion

The typical use case is to branch between an asymptotic and an exact method
depending on whether the criterion is satisfied. The example below estimates
the stratified difference in proportions with the CMH method when the criterion
holds, and with the unconditional exact method otherwise:

```{r}
is_mf_satisfied <- mantel_fleiss_crit(tbl)

if (is_mf_satisfied) {
  # Large enough sample: use the asymptotic CMH estimate.
  prop_diff_cmh(rsp, grp, strata)$diff
} else {
  # Sparse data: fall back to the exact (unstratified) method.
  prop_diff_uncond_exact(rsp, grp)$diff
}
```

The same idea can be used to select a test statistic. Here the CMH test is used
when the criterion holds, and Fisher's exact test on the collapsed table
otherwise:

```{r}
if (is_mf_satisfied) {
  prop_cmh(tbl)
} else {
  prop_fisher(table(grp, rsp))
}
```

## Empty strata

Strata that contain no observations carry no information and are dropped before
the statistic is computed. If *every* stratum is empty there is nothing to
compute, so the criterion is undefined and `mantel_fleiss_crit()` returns `NA`
(with an `NA` value attribute when `include_value = TRUE`):

```{r}
empty_tbl <- table(
  factor(character(0), levels = c("Active", "Control")),
  factor(logical(0), levels = c("TRUE", "FALSE")),
  factor(character(0), levels = "A")
)

mantel_fleiss_crit(empty_tbl, include_value = TRUE)
```

When branching on the result, remember to handle this `NA` case explicitly if
your data can produce fully empty tables.

## References
